Numerical simulations of the Cordilleran ice sheet through the last glacial cycle

### 2014-07-16 cordillera-cycle.tex
Work in progress on section 4.2.

@@

Although snapshots from the model output (Fig.~\ref{fig:snapshots}) give a
crisp picture of modelled ice dynamics at specific time frames, they often
can not be directly compared to geomorphological evidence, which is by nature
time-transgressive. In this section, we use the 6-km horizontal-resolution
model output from simulations driven by the GRIP and EPICA records to produce
a series of time-integrated plots containing information aggregated over the
whole simulation period.

During most of the glacial cycle, modelled ice cover is restricted to disjoint
ice caps centred on major mountain ranges of the North American Cordillera
(Fig.~\ref{fig:duration}). In both simulations driven by the GRIP and EPICA
records, a~contiguous ice cover spanning from the Alaska Range to the Rocky
Mountains exists for about 29\,kyr in total over MIS~2 and~4.
With the exception of the oceanic front and the northern foothills of the Alaska
Range, most of the maximal extent of the ice sheet is associated with short
durations of ice cover, generally a few thousand years, distributed over one
(GRIP) or two (EPICA) glaciations (Fig.~\ref{fig:duration}). This illustrates
that the most extensive stages of the modelled ice sheet are both short-lived
and out of balance with contemporary climate states, which confirms previous
inferences of a short-lived Cordilleran ice sheet based on the geological
record \citep{Clague.etal.1980, Stroeven.etal.2010}.

However, our simulations depict, independently of the choice of record used for
palaeo-temperature forcing, a central Skeena mountain ice cap persisting
throughout the entire last glacial cycle (Figs.~\ref{fig:snapshots}
and~\ref{fig:duration}). In particular, this ice cap persists during MIS~3,
providing a nucleation centre for the onset of glacial
readvance towards the last glacial maximum (MIS~2).

During time periods when a fully-grown Cordilleran ice sheet is absent from our
simulation, the presence of the Skeena ice cap can be explained by a more
widespread winter precipitation there than in other parts of the modelling
domain (Fig.~\ref{fig:atm}). Along most of the north-western North American
shore, coastal mountain ranges provide a sharp orographic barrier to westerly
winds, capturing atmospheric moisture in the form of precipitation, and leaving
interior lowlands arid. However, near the centre of our modelling domain, this
barrier is less pronounced than
elsewhere, allowing westerly winds to carry moisture further inland, until it
is progressively captured by the broad and mildly elevated group of the Skeena
Mountains.

In addition, the erosional landscape of the Skeena mountain bears one of the
strongest glacial imprint found within the area formerly covered by the
Cordilleran ice sheet \aref%
\footnote{\textbf{Arjen, Martin, Johan,} can we use a reference here to support
    this point (observations, mapping, shaded relief)? Alternatively, we could
    maybe use a photograph, if any of you has one. It would add a bit of
    variation into the figure list. \textbf{Johan}, you seemed to support this
    this point during our discussion, any idea?}.
Here we suggest that persistent ice cover (Fig.~\ref{fig:duration})
associated with basal ice temperatures at the pressure-melting point
(Fig.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}%
\footnote{Fig.~\ref{fig:erosion} is tricky because it may be significantly
    affected by the constant-geothermal heat flux assumption, which is not
    valid for this region. Arjen, Johan and I decided to leave it for my kappa
    for now. Regarding Figs.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}, I
    still wonder if it is useful to show both. Basically we have
    Fig.~\ref{fig:warmbase}\,=\,Fig.~\ref{fig:duration}\,*\,Fig.~\ref{fig:warmfrac}.
    Maybe the reader can perform this multiplication visually without need of
    Fig.~\ref{fig:warmbase}? Tell me what you think of it.})
may partly explain this spectacular landscape.

>>>>

\todo{Add general statements on the scale of glaciation (but where?).}

\todo{Put more equal weight to the northern sector}

@@

\todo{Mention the major nucleation centres. Growth of the CIS from the
      highlands.}

Although snapshots from the model output (Fig.~\ref{fig:snapshots}) yield a
clear spatial picture of modelled glaciation patterns at specific time frames,
they often can not be directly compared to geomorphological evidence, which is
by nature time-transgressive. In this section, we use high-resolution
simulations driven by the GRIP and EPICA records to locate the major nucleation
centres of the Cordilleran ice sheet and assess their longevity during the last
glacial cycle.

\todo{Rewrite this opening paragraph. Fig.~\ref{fig:snapshots} can actually be
      compared to geological evidence.}

During most of the glacial cycle, modelled ice cover is restricted to disjoint
ice caps centred on major mountain ranges of the North American Cordillera
(Fig.~\ref{fig:duration}). A 2\,500\,km-long continuous ice cover spanning from
the Alaska Range in the northeast to the Rocky Mountains in the southwest only
exists\renote{
    I chosed to write about model output in the present form to discern in from
    the reality.}
for at most 29\,kyr in total over the entire last glacial cycle. However,
except for its marine margin and the northern foothills of the Alaska Range,
the maximum extent of the ice sheet is attained for an even more brief period
of a few thousand years (Figs.~\ref{fig:timeseries}--{fig:duration}). This
result illustrates that the maximum extents of the modelled ice sheet during
MIS~4 and 2 were both short-lived and therefore out of balance with
contemporary climate, which confirms previous inferences of a short-lived
Cordilleran ice sheet based on the geological record \citep{Clague.etal.1980,
Stroeven.etal.2010}.

\todo{Include two paragraphs from Martin, but not here. Make sure that figures
      come in the right order in these paragraphs}.

Pehaps one of the most persistent landscape feature is the Skeena Mountains ice
cap, which persisted throughout the entire last glacial cycle
(Figs.~\ref{fig:snapshots} and~\ref{fig:duration}). Regardless of the applied
forcing (Fig.~\ref{fig:snapshots}, middle panels), this ice cap appears to
survive MIS~3, and serves as a nucleation centre at the onset of the glacial
readvance towards the LGM (MIS~2). The importance of residual ice for Noeth
American glacial history leading up to the LGM has been illustrated by the
MIS~3 residual ice bodies in northern and eastern Canada as nucleation centres
for a much more extensive MIS~2 configuration \citep{Kleman.etal.2010}.

The presence of a Skeena Mountains ice cap during most of the last glacial
cycle can be explained by a more widespread\renote{
    I mean widespread, precipitation there is lower than elsewhere.},
current winter precipitation for that region there than for other parts of the
modelling domain (Fig.~\ref{fig:atm}). Along most of the north-western coast of
North America, coastal mountain ranges form a pronounced topographic barrier
for westerly winds, capturing atmospheric moisture in the form of orographic
precipitation, and resulting in arid interior lowlands. However, near the
centre of our modelling domain, this barrier is less pronounced than elsewhere,
allowing westerly winds to carry moisture\renote{
    I do not agree with ``carry precipitation''. When reading this I get the
    picture of water drops falling diagonaly under action of the wind. I do not
    deny that transport of precipitation happens and
    is an important process for understanding orographic
    precipitation (think of why it is raining so much in the Swedish mountains
    even when the wind comes from the west), but it is not what I am trying to
    describe here. I wrote ``moisture'' and this is what I mean.}
further inland, until it is progressively captured by the extensive, but mildly
elevated group of the Skeena Mountains in north-central British Columbia.

\todo{Explain why it would not be good with general statements on thermal
      evolution of ice cover.}

[In addition, the erosional landscape of the Skeena mountain bears one of the
strongest glacial imprint found within the area formerly covered by the
Cordilleran ice sheet \aref%
\footnote{\textbf{Arjen, Martin, Johan,} can we use a reference here to support
    this point (observations, mapping, shaded relief)? Alternatively, we could
    maybe use a photograph, if any of you has one. It would add a bit of
    variation into the figure list. \textbf{Johan}, you seemed to support this
    this point during our discussion, any idea?}.
Here we suggest that persistent ice cover (Fig.~\ref{fig:duration})
associated with basal ice temperatures at the pressure-melting point
(Fig.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}%
\footnote{Fig.~\ref{fig:erosion} is tricky because it may be significantly
    affected by the constant-geothermal heat flux assumption, which is not
    valid for this region. Arjen, Johan and I decided to leave it for my kappa
    for now. Regarding Figs.~\ref{fig:warmfrac} and ~\ref{fig:warmbase}, I
    still wonder if it is useful to show both. Basically we have
    Fig.~\ref{fig:warmbase}\,=\,Fig.~\ref{fig:duration}\,*\,Fig.~\ref{fig:warmfrac}.
    Maybe the reader can perform this multiplication visually without need of
    Fig.~\ref{fig:warmbase}? Tell me what you think of it.})
may partly explain this spectacular landscape.]

A correlation is observed between the modelled duration of warm based ice cover
(Fig.~\ref{fig:warmbase}) and the degree of glacial modification of the
landscape (mainly in terms of the development of deep glacial valleys and
troughs). We find evidence for this on both the southwestern and northeastern
slopes of the southern Coast Mountains, on the western slopes of the Columbia
Mountains, on the western and eastern slopes of the northern Coast Mountains
and the Saint-Elias Mountains, and radiating off the Skeena Mountains
(Figs.~\ref{fig:duration}, \ref{fig:warmbase} and~X\renote{
    Yes it could be interesting to compare basal flow patterns with Fig.~2 of
    \citet{Kleman.etal.2010, but I don't think that I have the time for this
    right now.}}).
The Skeena Mountains, for example, indeed bear a strong glacial imprint that
indicates ice drainage in a system of distinct glacial troughs emanating in a
radial pattern from the centre of the mountain range (Fig. X). We suggest that
persistent ice cover (Fig.~\ref{fig:duration}) associated with basal ice
temperatures at the pressure-melting point (Figs.~\ref{fig:warmfrac}
and~\ref{fig:warmbase}) explains the large-scale glacial erosional imprint on
the landscape. In contrast, a well-developed network of glacial valleys west of
the MacKenzie Mountains \citep[Fig.~X;][]{Kleman.etal.2010,
Stroeven.etal.2010}, which is modelled to have hosted warm-based ice
(Fig.~\ref{fig:warmfrac}) has only been glaciated for a short duration
(Fig.~\ref{fig:duration}) during the last glacial cycle according to our
results (Figs.~\ref{fig:duration} and~\ref{fig:warmbase}). We therefore infer
that either our model does not perform reliably in this part of the domain or
the observed landscape pattern originates from older glacial cycles and
indicates an increased relative importance of this ice dispersal centre prior
to the Late Pleistocene \citep[cf.][]{Ward.etal.2008, Demuro.etal.2012}.

\todo{Paragraph on model uncertainties related to warm-based ice.}
